---
title: "US imports of Goods from World"
author: "Monica Boghean"
date: "July, 20, 2026"
output:
  html_document:
    toc: true
    toc_depth: 3
    number_sections: true
    df_print: paged
editor_options: 
  markdown: 
    wrap: sentence
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  warning = FALSE,
  message = FALSE,
  fig.width = 10,
  fig.height = 6
)
```

# Abstract

This project applies modern time series methods to monthly U.S. imports of goods (FRED series IMP0015) from 2000–2026 on a customs basis. Customs‑basis valuation captures the goods’ value plus certain freight and insurance costs up to the U.S. port of import, but excludes import duties and domestic charges.
The series is transformed to the natural logarithm to stabilize variance and then subjected to unit‑root testing (KPSS, ndiffs, nsdiffs) and systematic differencing to obtain a stationary representation. Using the tsibble, feasts, and fable frameworks in R, I analyze the autocorrelation structure via ACF/PACF plots and estimate a range of seasonal ARIMA and ETS models, alongside naive and seasonal naive benchmarks. Model selection is based on information criteria (AIC, AICc, BIC), residual diagnostics (ACF of residuals, Ljung–Box tests), and out‑of‑sample forecast accuracy on a held‑out test set. The results demonstrate that, despite the presence of a well‑fitting seasonal ARIMA model, a simple seasonal naive specification on the log scale achieves the lowest RMSE, MAE, and MAPE, underscoring the strength of annual seasonality in U.S. imports and the competitiveness of benchmark forecasting methods.

Objectives:

1. Apply log transformation and formal unit‑root tests to assess stationarity and determine appropriate non‑seasonal and seasonal differencing orders for monthly import data.
2. Use ACF/PACF diagnostics on differenced series to guide the specification of candidate seasonal ARIMA models within the fable framework.
3. Estimate and compare automatic ARIMA, ETS, naive, and seasonal naive models using information criteria, residual ACF, and Ljung–Box tests.
4. Evaluate forecast performance on a chronologically held‑out test set and identify the model that minimizes RMSE, MAE, MAPE on both log and back‑transformed scales.

# Load packages

```{r packages}
library(tidyverse)
library(lubridate)
library(tsibble)
library(feasts)
library(fable)
library(fabletools)
library(readr)
library(scales)
library(dplyr)

```

# Data import

```{r data-import}
imports_raw <- read_csv("IMP0015.csv")
glimpse(imports_raw)
head(imports_raw)

```

# Data preparation

```{r data-prep}
imports_ts <- imports_raw %>%
  mutate(observation_date = yearmonth(observation_date)) %>%
  filter(observation_date >= yearmonth("2000-01")) %>%
  as_tsibble(index = observation_date)

imports_ts
```

Key dataset features:

-   Dataset source: https://fred.stlouisfed.org/series/IMP0015.
-   The FRED dataset is a time series containing information about the imports of goods by customs basis from all over the world, measured in millions of dollars. The series is not seasonally adjusted.
-   Frequency: Monthly.
-   Period covered in the analysis: original series (1987-01-01 to 2026-05-01), restricted for this analysis to 2000-01-01 to 2026-05-01.
-   The series captures strong upward long-run trend, reflecting globalization, economic growth, and price changes, along with sharp declines due to global financial crisis and COVID-19 shock, followed by a rapid rebound.

# Time plot

```{r time-plot, fig.cap="Monthly imports of goods from all the countries"}
imports_ts %>%
  autoplot(IMP0015) +
  labs(
    title = "Monthly Imports of Goods Time Series",
    x = "Date",
    y = "Imports of Goods (millions of dollars)"
  ) +
  theme_minimal()
```

Main insights from the time plot: 
- The series has an upward trend from 2000s to early 2026s, with import levels rising substantially from around 90,000-100,000 million dollars in the early 2000s to above 300,000 million dollars in the most recent years. Although the growth is not constant (presumably due to changing macroeconomic conditions and trade policies over time),the long-run direction is strongly positive, especially after 2020.
- The monthly pattern exhibits regular seasonal fluctuations, and the amplitude of seasonal swings appears to increase as the level of the series rises, suggesting that seasonal effects scale with overall imports.
- One pronounced trough is before 2010, followed by strong rebound, consistent with a major global shock and subsequent recovery in trade flows. Also, a similar sharp decline appears around the early 2020s, followed again by strong growth, indicating that imports are sensitive to global disruptions but tend to revert quickly once conditions normalize.

# Log transformation of the series - since variance fluctuates with the level of the series

```{r log-transform}
imports_log <- imports_ts %>%
  mutate(log_IMP0015 = log(IMP0015))

# quick check
head(imports_log)
```

Log‑scale time plot:

```{r log-plot, fig.cap="Monthly imports of goods on a log scale"}
imports_log %>%
  autoplot(log_IMP0015) +
  labs(
    title = "Monthly Imports of Goods (log scale)",
    y = "log(IMP0015)"
  )
```
After applying a logarithmic transformation to monthly imports, the variance appears approximately stabilized across time. The amplitude of month‑to‑month movements is now relatively constant, indicating that the log transformation has reduced the heteroscedasticity observed in the original series. On this scale, the remaining non‑stationarity is primarily due to a persistent upward trend and structural shocks (for example, the sharp decline around the late 2000s and early 2020s), which can be addressed through differencing and appropriate ARIMA or ETS modeling.
All subsequent differencing (non‑seasonal and seasonal) will be performed on the log‑transformed series, so that the resulting differenced series represents relative (percentage) changes rather than absolute changes.

# Stationarity and differencing

## KPSS test on the log transformed series

```{r kpss-original}
imports_log %>%
  features(log_IMP0015, unitroot_kpss)
```

The KPSS test uses a null hypothesis that the time series is stationary (either around a fixed mean or, in the trend version of the test, around a deterministic trend), and an alternative hypothesis that the series is non‑stationary and contains a unit root.
In our application, the p‑value of 0.01 leads us to reject the null, indicating that the original import series is non‑stationary and requires first differencing to achieve stationarity.

## First difference

```{r first-difference}
imports_diff <- imports_log %>%
  mutate(diff_1 = difference(log_IMP0015))

imports_diff %>%
  autoplot(diff_1) +
  labs(
    title = "First Difference of the Log Transformed Series",
    x = "Date",
    y = "First difference"
  ) +
  theme_minimal()
```

After applying the first difference to the log transformed import series, the differenced values fluctuate around a relatively stable mean with no clear long‑run trend, while preserving the regular seasonal pattern and the large shocks around the early 2010s, 2020s and 2026s.
Visually, the series looks much more like a stationary process than the original level series.

```{r kpss-diff1}
imports_diff %>%
  features(diff_1, unitroot_kpss)
```

To formalize this, I ran the KPSS test on the first‑differenced log transformed series.
The KPSS test has a null hypothesis of stationarity and an alternative hypothesis of non‑stationarity; in this case the p‑value is 0.10, which is above conventional significance levels.
Therefore, we fail to reject the null hypothesis and treat the first‑differenced log transformed series as stationary.
This result, combined with the visual inspection, indicates that a single regular difference (d=1) is sufficient on the log transformed data, and additional differencing is not required before fitting ARIMA models.

However, I will continue the analysis by doing also seasonal and double differences and see the impact on the time series.

## Seasonal difference

```{r seasonal-difference}
imports_seas <- imports_log %>%
  mutate(seas_diff = difference(log_IMP0015, 12))

imports_seas %>%
  autoplot(seas_diff) +
  labs(
    title = "Seasonal Difference of the Log Transformed Series",
    x = "Date",
    y = "Seasonal difference"
  ) +
  theme_minimal()
```

```{r kpss-seas}
imports_seas %>%
  features(seas_diff, unitroot_kpss)
```

After computing the seasonal difference with lag 12, the log series doesn't seem to remain around a constant mean.
Visually, the seasonal-differenced series looks more volatile and less stable around a mean, but the KPSS test still fails to reject stationarity.

However, I will go further and do a double difference of the log time series and analyze results.

## Double difference

```{r double-difference}
imports_double <- imports_log %>%
  mutate(double_diff = difference(difference(log_IMP0015), 12))

imports_double %>%
  autoplot(double_diff) +
  labs(
    title = "Double Difference of the Log Transformed Series",
    x = "Date",
    y = "Double difference"
  ) +
  theme_minimal()
```

```{r kpss-double}
imports_double %>%
  features(double_diff, unitroot_kpss)
```

Finally, I computed a "double difference" by applying both the regular and seasonal differences to the log transformed series.
The resulting plot shows a series that is almost completely flat around zero, with most variation concentrated in a few large spikes.
This indicates that much of the underlying structure has been removed, leaving a highly differenced process with limited remaining signal.
The KPSS test on the double‑differenced series yields a p‑value of 0.10, so we again fail to reject the null hypothesis of stationarity.
However, given that the first difference alone already produces a stationary series on the log transformed series, the additional differencing is not necessary for stationarity and instead tends to over‑difference the data.
Over‑differencing can introduce artificial negative autocorrelation and inflate noise, making the series harder to model and interpret.

## Required differencing

```{r ndiffs-check}
imports_log %>%
  features(log_IMP0015, unitroot_nsdiffs)

imports_log %>%
  features(log_IMP0015, unitroot_ndiffs)
```

To cross-check the standard procedures regarding stationarity analysis, I used the nsdiffs() and ndiffs() functions from the forecast/fable framework.
These functions apply seasonal and non‑seasonal unit‑root tests and return the minimum number of differences required to obtain a stationary series.
For this log transformed imports series, nsdiffs() returns 0, indicating that no seasonal differencing is required at the monthly frequency.
In contrast, ndiffs() returns 1, which implies that one regular difference is sufficient to remove the unit root and achieve stationarity.
This aligns with the KPSS results: the level of the log series is non‑stationary, while the first‑differenced of the log transformed series has a p‑value of 0.10 and appears stationary.
Following the standard recommendation to use the smallest differencing orders that satisfy stationarity, I set the ARIMA differencing parameters to d=1 and D=0.
All subsequent model identification, estimation, and forecasting are therefore based on a series that has been under a log transformation, differenced once non‑seasonally, without additional seasonal differencing.

However, for learning purposes, I will look at the ACF and PACF plots for the first differenced log series, seasonal differenced log series and double differenced log series.

# ACF and PACF

## First-differenced log series

```{r acf-pacf-diff1, fig.height=8}
imports_diff %>%
  gg_tsdisplay(diff_1, plot_type = "partial", lag_max = 36) +
  labs(title = "First-Differenced Log Transformed Series with ACF and PACF")
```

Applying a logarithmic transformation to imports stabilized the variance but did not materially impacted the autocorrelation and partial autocorrelation structure of the first‑differenced series. This indicates that the underlying dynamic relationships across months may have similarities on both the original and log scales, while the log transformation primarily improves variance properties and interpretability (differences as growth rates).
As such, the top panel shows the first difference of the log transformed time series fluctuating around a roughly constant 0 mean with no visible trend, preserving a clear seasonal pattern.
There are a few large spikes that likely correspond to major economic shocks, but outside those episodes, the variance looks relatively stable over time.
The ACF shows many bars outside the significance bounds at many lags, not just at the seasonal lags. This suggests that there is still meaningful autocorrelation in the differenced log series, so it is not behaving like white noise.
The PACF also shows a prominent spike at lag 12, with smaller spikes at subsequent seasonal lags.
This indicates that including at least one seasonal autoregressive term at the yearly lag is appropriate.

## Seasonal-differenced series

```{r acf-pacf-seas, fig.height=8}
imports_seas %>%
  gg_tsdisplay(seas_diff, plot_type = "partial", lag_max = 36) +
  labs(title = "Seasonal-Differenced Log Series with ACF and PACF")
```

After doing a seasonal difference on the log series, we can observe that the series oscilates around zero, with clear cycles but no obvious trend, suggesting that the combination of log transformation and seasonal differencing has largely removed both level and seasonality. Large swings around the late 2000s and early 2020s remain, reflecting structural shocks rather than regular seasonality.
With regards to the ACF, the figure shows a slow, geometric-like decay from high positive values at low lags crossing zero and becoming negative around lags 10-30, before tapering back towards zero.
The PACF on the other side has a large spike at lag 1, followed by smaller spikes at subsequent lags, most of them falling within the significance bounds. 

## Double-differenced series

```{r acf-pacf-double, fig.height=8}
imports_double %>%
  gg_tsdisplay(double_diff, plot_type = "partial", lag_max = 36) +
  labs(title = "Double-Differenced Log Series with ACF and PACF")
```

The double difference combines the regular and seasonal differences on the log series, producing a series that is almost completely flat around zero, with most variation concentrated in a few very large spikes around the structural shocks.
This indicates that much of the underlying trend and seasonal structure has been stripped out, leaving a highly differenced process dominated by outliers rather than by smooth dynamics.
The ACF shows a lot of structure, with many positive and negative spikes outside the blue significance bands. This indicates strong remaining autocorrelation, so the double differenced log series still has meaningful dependence across time rather than behaving like white noise. 
The PACF has a positive noticeable spike at low lag and a pronounced negative spike at lag 12, while most other lags lie within the significance bands. Overall, the PACF looks more regular than the ACF, suggesting that a low-order AR structure could summarize much of the remaining dependence.
Although the KPSS p‑value remains at 0.10 (so we fail to reject stationarity), the combination of an almost flat series and pronounced negative autocorrelation suggests that the second round of differencing introduces artificial dynamics rather than revealing useful information.
In line with the recommendation to use the smallest differencing order that achieves stationarity, I therefore treat the double difference as excessive and retain only one non‑seasonal difference, setting d=1 and D=0 for the ARIMA modeling. This keeps the model parsimonious while avoiding the distortions associated with over-differencing.

# ARIMA modeling

## Candidate models

```{r arima-models}
fit_arima <- imports_log %>%
  model(
    arima_1 = ARIMA(log_IMP0015 ~ pdq(1,1,0) + PDQ(0,0,3)),
    auto_arima = ARIMA(log_IMP0015, stepwise = FALSE, approx = FALSE)
  )

fit_arima
```

Explanation of why this candidate model was considered:

\<ARIMA(1,1,0)(0,0,3)[12]\>

 * In the PACF figure of the first differenced log series, up to lag 12, the largest spike is the negative value at lag 1, which corresponds to a non-seasonal AR(1) term.
 * Only one regular difference was applied to the log transformed series, so the non-seasonal differencing order is d=1 and the seasonal differencing order is D=0.
 * In the ACF figure, there are significant positive spikes at lags 12, 24 and 36, which correspond to a seasonal MA(3) component with period 12.

## Information criteria

```{r arima-ic}
arima_ic <- fit_arima %>%
  glance() %>%
  select(.model, AICc, AIC, BIC)

arima_ic
```

To compare the candidate ARIMA specifications, I extracted the AICc, AIC, and BIC values for each model using glance().
The AIC, AICc, and BIC values are negative because the fitted ARIMA models achieve high likelihoods relative to their parameter counts. Since these criteria are on a relative scale, the absolute sign is not important; the more negative values for the auto_arima specification indicate a better trade‑off between goodness of fit and model complexity than for the manually specified ARIMA model.
Therefore, the best arima model is the one with the lowest AICc, in this case the auto arima model.

# Training and test split

To compare the out-of-sample forecasting performance of the selected ARIMA model and an ETS benchmark, I split the log series chronologically into training and test sets.
For log time series data, this split must preserve the temporal ordering of the observations, so the earlier observations are used for model estimation and the later observations are held out for forecast evaluation.

```{r train-test-split}
train <- imports_log %>%
  filter(observation_date <= yearmonth("2022 Dec"))

test <- imports_log %>%
  filter(observation_date >= yearmonth("2023 Jan"))

train
test
```

In this setup, the training set runs from the beginning of the sample through December 2022, while the test set covers January 2023 through the end of the available log series.
This creates a genuine out-of-sample evaluation period for comparing forecast accuracy.

# Fit ARIMA and ETS on the training set

```{r fit-train-models}
fit_train <- train %>%
  model(
    auto_arima = ARIMA(log_IMP0015, stepwise = FALSE, approx = FALSE),
    ets_auto = ETS(log_IMP0015)
  )

fit_train
```

Using the log‑transformed imports series in the training set, I estimated two automatic forecasting models:

* An automatic seasonal ARIMA model, selected by the ARIMA() function, with specification ARIMA(3,1,0)(2,0,0)[12].This model includes one regular difference on the log scale, three non‑seasonal autoregressive terms, and two seasonal autoregressive terms with annual seasonality.
* An automatic ETS model, denoted ETS(A, Ad, A), which is an exponential smoothing model with additive errors, damped additive trend, and additive seasonality.

These models are fitted to the log‑transformed training data, and will be evaluated on a held‑out test set. I will compare their in‑sample specifications and residual diagnostics, then generate forecasts over the test horizon and assess performance using RMSE, MAE, and MAPE (on both the log scale and the original imports scale after back‑transformation) to determine which model delivers better out‑of‑sample accuracy.

# ARIMA output

```{r arima-train-report}
fit_train %>%
  select(auto_arima) %>%
  report()
```

The automatic ARIMA selection on the training data yields a seasonal ARIMA model of the form ARIMA(3,1,0)(2,0,0)[12] for the log transformed imports series.
This specification uses one non‑seasonal difference to remove the unit root, three non‑seasonal autoregressive terms AR(3), and two seasonal autoregressive terms, AR(2).
The estimated non‑seasonal AR coefficients are -0.2581, 0.0708, and 0.393 with standard errors of 0.0577, 0.0589, and 0.058 respectively. The seasonal AR coefficients are 0.4975, and 0.2611, with standard errors of 0.0603, and 0.0641, indicating that the seasonal AR effects are statistically significant and play an important role in capturing the yearly pattern in the differenced log series.
The innovation variance is estimated at approximately 0.002036, and the model attains an AIC of -907.35, AICc of -907.04, and BIC of -885.65 on the training set.


# ARIMA residual diagnostics

```{r arima-train-diagnostics, fig.height=8}
fit_train %>%
  select(auto_arima) %>%
  gg_tsresiduals(lag=36)
```

The residual diagnostic plots clearly indicate: 
1. Plot 1 (Innovation Residuals) – data is mainly flat and stationary, moving around a mean of 0 - no obvious remaining trend or volatility.
It shows some large negative spikes before January 2010 and after January 2020, suggesting some major structural shocks in the data.
2. Plot 2 (ACF plot) – several spikes extending beyond the blue significance bands, both positive and negative, meaning that there is leftover structure the ARIMA model hasn't fully explained.
3. Plot 3 (Residual Histogram) – residuals look roughly symmetric and bell-shaped, with many values near zero, suggestiong the Gaussian error assumption is broadly reasonable.

So, residual diagnostics indicate that, while the ARIMA(3,1,0)(2,0,0)[12] model removes most of the systematic structure in the log‑differenced imports series, the residual ACF still exhibits several significant spikes. This suggests remaining short‑run dependence, so alternative ARIMA specifications with additional or adjusted AR/MA terms will be considered and compared later in this project based on information criteria, residual autocorrelation, and out‑of‑sample accuracy.

But first, I will also do the Ljung-Box test to see if the residuals are independent (white noise) or correlated.

```{r arima-train-ljung}
fit_train %>%
  select(auto_arima) %>%
  augment() %>%
  features(.innov, ljung_box, lag = 36, dof = 5)
```
A Ljung–Box test applied to the ARIMA residuals up to lag 36 yields a test statistic of 58.81 and a p‑value of 0.0018, indicating significant remaining autocorrelation. Consequently, the residuals do not behave as white noise, and the ARIMA(3,1,0)(2,0,0)[12] specification is not fully adequate in capturing all dynamic structure in the log‑transformed imports series.


Next, I will fit an ETS model to the training dataset.

# ETS output

```{r ets-train-report}
fit_train %>%
  select(ets_auto) %>%
  report()
```

Check the residual diagnostics for the ETS model:

# ETS residual diagnostics

```{r ets-train-diagnostics, fig.height=8}
fit_train %>%
  select(ets_auto) %>%
  gg_tsresiduals(lag=36)
```

Residual diagnostics for the ETS(A, Ad, A) model on the log‑transformed imports series indicate that the model does not adequately capture the temporal dependence. The residual ACF displays numerous significant spikes across many lags, suggesting pronounced remaining autocorrelation. At this time, this evidence supports favoring the ARIMA specification over ETS for this series. However, the models will be compared by their specifications on a hold-out test set.

```{r ets-train-ljung}
fit_train %>%
  select(ets_auto) %>%
  augment() %>%
  features(.innov, ljung_box, lag = 36, dof = 4)
```

For the ETS(A, Ad, A) model, the Ljung–Box test was computed up to lag 36 with degrees of freedom adjusted for the four smoothing parameters (alpha, beta, gamma, and phi). The resulting test statistic is 243.29 with a p‑value effectively equal to 0, far below conventional significance levels. This confirms that substantial autocorrelation remains in the residuals, reinforcing the conclusion that the ETS specification does not adequately capture the dynamics of log‑transformed imports.


# Summary so far
So far, my diagnostics say:
* ARIMA fits much better than ETS by exploring the residual diagnostics plots.
* ARIMA residuals have less autocorrelation than ETS, but still some remaining structure and a low Ljung-Box p-value.
* The comparison between ARIMA and ETS will be done on a hold-out test set, to have a clear picture in terms of RMSE, MAE, and MAPE values.

Going further, I will try a few alternative ARIMA models at the price that a more complex model might reduce residual autocorrelation but lose simplicity at the same time. This is why I will also include two simpler benchmarks: naive and seasonal naive models.


# Next steps - adjust ARIMA models to obtain an algorithm with better performance on the log series and add the naive and snaive models to the log transformed series

```{r arima-models}
fit_arima <- imports_log %>%
  model(
    auto_arima = ARIMA(log_IMP0015, stepwise = FALSE, approx = FALSE),
    arima_2 = ARIMA(log_IMP0015 ~ pdq(3,1,0) + PDQ(0,0,3)),
    arima_3 = ARIMA(log_IMP0015 ~ pdq(2,1,1) + PDQ(2,0,0)),
    arima_4 = ARIMA(log_IMP0015 ~ pdq(2,1,1) + PDQ(0,0,2))
    )

fit_arima
```
```{r arima-ic}
arima_ic <- fit_arima %>%
  glance() %>%
  select(.model, AICc, AIC, BIC)

arima_ic
```

Besides the automatic ARIMA(3,1,0)(2,0,0)[12] model, I also estimated several alternative ARIMA specifications (arima_2, arima_3, and arima_4) that vary the non‑seasonal AR/MA orders and the seasonal components. These additional models were compared using AIC, AICc, and BIC, but none achieved lower information‑criterion values than the automatic ARIMA specification on the full log series. In the subsequent train/test forecast comparison, I therefore focus on the automatic ARIMA model, an ETS benchmark, and simple naive and seasonal naive models, to see which of these models will provide the best out-of-sample accuracy.


```{r fit-train-models}
fit_train <- train %>%
  model(
    auto_arima = ARIMA(log_IMP0015, stepwise = FALSE, approx = FALSE),
    ets_auto   = ETS(log_IMP0015),
    naive_log  = NAIVE(log_IMP0015),
    snaive_log = SNAIVE(log_IMP0015)
    )

fit_train
```

# Forecast comparison on the test set

```{r forecast-test}
h_test <- nrow(test)
h_test

fc_test <- fit_train %>%
  forecast(h = h_test)

fc_test
```

```{r accuracy-test-on log series}
acc_compare <- fc_test %>%
  accuracy(test) %>%
  select(.model, RMSE, MAE, MAPE, MASE)

acc_compare
```

Forecast accuracy on the held‑out test set was evaluated for all models on the log‑transformed imports series. The seasonal naive model (SNAIVE), which simply repeats last year’s log imports for each month, achieved the lowest RMSE (0.084), MAE (0.066) and MAPE (0.52), outperforming the automatic ARIMA, ETS, and non‑seasonal naive models. This indicates that, given the strong and stable annual seasonality in monthly imports, the seasonal naive method provides the most accurate and parsimonious forecasts over the assessment horizon. In contrast, ETS is the model that has the worst performance in terms of RMSE, MAE, and MAPE on the hold-out test set.

# Forecast plots against actual values on the log transformed series

```{r fc-plot-all-log}
fc_test %>%
  autoplot(bind_rows(train, test)) +
  autolayer(test, log_IMP0015, color = "black") +
  labs(
    title = "All Models Forecasts vs Actual Values on the log transformed series",
    x = "Date",
    y = "Imports"
  ) +
  theme_minimal()
```


# Best model on the log tranformed series 

To identify an appropriate forecasting model for monthly imports, I first transformed the series using the natural logarithm to stabilize the variance and then applied non‑seasonal differencing to remove the stochastic trend. On this log‑differenced series, I explored a range of seasonal ARIMA and ETS specifications, guided by ACF/PACF plots, and residual diagnostics (ACF of residuals and Ljung–Box tests). These diagnostics showed that, at a first glance, ETS(A, Ad, A) provided a relatively poor in‑sample fit with strong remaining autocorrelation in the residuals, while the automatic seasonal ARIMA model achieved a much better fit but still left some residual dependence at longer lags.
Recognizing that simple benchmarks are often competitive in practice, I then fitted two baseline models on the same log‑transformed training data: a non‑seasonal naive model and a seasonal naive model that repeats last year’s log imports for each month. All models, ARIMA, ETS, naive, and seasonal naive, were evaluated on a held‑out test set using RMSE, MAE, and MAPE on the log scale. The seasonal naive model attained the lowest RMSE, MAE, and MAPE among all methods, with the naive model second, and both ARIMA and ETS performing worse over the forecast horizon.
Taken together, these results indicate that the log‑scale seasonal naive model is the most accurate and parsimonious choice for this series. It leverages the strong and stable annual seasonality present in monthly imports, avoids the complexity and residual autocorrelation issues of more elaborate ARIMA and ETS specifications, and provides the best out‑of‑sample forecast accuracy relative to all other models considered. Consequently, the final forecasts in this study are based on the seasonal naive model, with confidence intervals reflecting the variability observed in the historical data.


# Model verification on the back transformed data

When converting the log‑scale forecasts back to the original imports scale by exponentiation, I implicitly assume that the forecast errors on the log scale are approximately symmetric and normal; this back‑transformation can introduce a small bias, and more sophisticated bias‑correction methods are available, but they are not necessary for the purposes of this study.


# Create original scale versions

```{r - Back-transform forecasts}
fc_test_orig <- fc_test %>%
  mutate(.mean = exp(.mean))

```
```{r - Back-transform actuals}
test_orig <- test %>%
  mutate(IMP0015 = exp(log_IMP0015))
  
```

# Compute accuracy on the original scale

```{r - accuracy measures on the original scale}
acc_orig <- fc_test_orig %>%
  accuracy(test_orig) %>%
  select(.model, RMSE, MAE, MAPE, MASE)

acc_orig

```
To validate that the model ranking is robust on the original imports scale, I back‑transformed both forecasts and actuals by exponentiating the log series. Forecast accuracy was then recomputed in the original units of monthly imports. The seasonal naive model continued to achieve the lowest RMSE, MAE, and MAPE among all models, confirming that its superior performance on the log scale translates into more accurate forecasts on the original scale as well. Naive random walk forecasts were second best, while ARIMA and ETS produced noticeably larger errors. This robustness across both scales strengthens the case for adopting the seasonal naive model as the final forecasting method.
Therefore, the models were estimated on the log‑transformed imports series to stabilize variance and improve stationarity properties. To assess forecast accuracy on the original imports scale without refitting separate models, the log‑scale forecasts and test observations were exponentiated and evaluated using RMSE, MAE, and MAPE in the original units. This approach preserves the benefits of modeling on the log scale while providing interpretable accuracy measures for the original data.


# Conclusion

This analysis shows that monthly U.S. imports of goods exhibit a strong long-run upward trend, clear annual seasonality, and noticeable disruptions associated with major economic shocks. After applying a logarithmic transformation to stabilize the variance, the series required one non-seasonal difference but no seasonal difference, thus d=1 and D=0 were appropriate for ARIMA modeling.
Several forecasting approaches were then evaluated, including automatic ARIMA and ETS, naive and seasonal naive models. Although the automatic ARIMA model provided the strongest in-sample fit, its residual diagnostics still indicated some remaining autocorrelation, while ETS model performed substantially worse by both information criteria and residual behavior.
When forecast performance was assessed on a held-out test set, the seasonal naive model achieved the lowest RMSE, MAE, and MAPE on both the log scale and the back transformed original scale. This result suggests that the dominant feature of the series is stable yearly seasonality, and that a simple benchmark model can be more effective for short-term forecasting than complex ARIMA or ETS alternatives in this case.
Overall, the project highlights an important forecasting principle: more sophisticated models do not always produce better predictions. For this specific dataset, the seasonal naive model offers the most accurate and parsimonious solution, while future work could explore bias-corrected back-transformation, intervention modeling for major shocks, or dynamic regression with external macroeconomic predictors to improve performance further.








